Legacy Data Confounds Genomics Studies

### 2019-08-15 LatexManuscript/MBE/legacy.tex
edited the discussion

An enrichment in mutational profiles in one chromosome (Figure \ref{CountPerChrom}B) is difficult to explain biologically or technologically as neither are expected to produce chromosome-specific differences in error rates. 
Bioinformatics data processing errors are one possibility that would be consistent with this as chromosomes are commonly treated and analyzed separately.
However, to pinpoint the precise technical source of the discrepancy would require further forensic inquiries into the details of the heterogeneous sample preparation and data processing pipelines used throughout the 1kGP. 
Given the progress in sequencing and calling that occurred since the early phases of the 1kGP (Figure \ref{MapQual}), it is likely that the source of these biases is not longer being actively introduced in recent sequence data.

@@

\bibliography{Legacy}

@@

\extrafloats{100}

>>>>

An enrichment in mutational profiles in one chromosome (Figure \ref{CountPerChrom}B) is difficult to explain through biological mechanisms or sequencing technology as neither are expected to produce systematic differences across chromosomes. 
Issues arising from subtle differences in how these data were integrated or processed are one explanation that would be consistent with this as chromosomes are commonly treated and analyzed separately.
However, to pinpoint the precise technical source of the discrepancy would require further forensic inquiries into the details of the heterogeneous sample preparation, combination of sequencing technologies and data processing pipelines used throughout the 1kGP. 
Given the progress in sequencing and calling that occurred since the early phases of the 1kGP (Figure \ref{MapQual}), it is likely that the source of these biases is not longer being actively introduced in recent sequence data.

@@

%\bibliography{Legacy}
\begin{thebibliography}{}

\bibitem[{1000 Genomes Project Consortium},
  2010]{1000GenomesProjectConsortium2010}
{1000 Genomes Project Consortium} (2010).
\newblock {A map of human genome variation from population-scale sequencing.}
\newblock {\em Nature}, 467(7319):1061--73.

\bibitem[{1000 Genomes Project Consortium},
  2012]{The1000GenomesProjectConsortium2012}
{1000 Genomes Project Consortium} (2012).
\newblock {An integrated map of genetic variation}.
\newblock {\em Nature}, 135:0--9.

\bibitem[Aikens et~al., 2019]{Aikens2018}
Aikens, R.~C., Johnson, K.~E., and Voight, B.~F. (2019).
\newblock {Signals of Variation in Human Mutation Rate at Multiple Levels of
  Sequence Context}.
\newblock {\em Molecular Biology and Evolution}.

\bibitem[Alexandrov et~al., 2013]{Alexandrov2013}
Alexandrov, L.~B., Nik-Zainal, S., Wedge, D.~C., Aparicio, S.~A., Behjati, S.,
  Biankin, A.~V., Bignell, G.~R., Bolli, N., Borg, A., B{\o}rresen-Dale, A.~L.,
  Boyault, S., Burkhardt, B., Butler, A.~P., Caldas, C., Davies, H.~R.,
  Desmedt, C., Eils, R., Eyfj{\"{o}}rd, J.~E., Foekens, J.~A., Greaves, M.,
  Hosoda, F., Hutter, B., Ilicic, T., Imbeaud, S., Imielinsk, M., J{\"{a}}ger,
  N., Jones, D.~T., Jonas, D., Knappskog, S., Koo, M., Lakhani, S.~R.,
  L{\'{o}}pez-Ot{\'{i}}n, C., Martin, S., Munshi, N.~C., Nakamura, H.,
  Northcott, P.~A., Pajic, M., Papaemmanuil, E., Paradiso, A., Pearson, J.~V.,
  Puente, X.~S., Raine, K., Ramakrishna, M., Richardson, A.~L., Richter, J.,
  Rosenstiel, P., Schlesner, M., Schumacher, T.~N., Span, P.~N., Teague, J.~W.,
  Totoki, Y., Tutt, A.~N., Vald{\'{e}}s-Mas, R., {Van Buuren}, M.~M., {Van 'T
  Veer}, L., Vincent-Salomon, A., Waddell, N., Yates, L.~R., Zucman-Rossi, J.,
  {Andrew Futreal}, P., McDermott, U., Lichter, P., Meyerson, M., Grimmond,
  S.~M., Siebert, R., Campo, E., Shibata, T., Pfister, S.~M., Campbell, P.~J.,
  and Stratton, M.~R. (2013).
\newblock {Signatures of mutational processes in human cancer}.
\newblock {\em Nature}, 500(7463):415--421.

\bibitem[Astle et~al., 2016]{astle2016allelic}
Astle, W.~J., Elding, H., Jiang, T., Allen, D., Ruklisa, D., Mann, A.~L., Mead,
  D., Bouman, H., Riveros-Mckay, F., Kostadima, M.~A., et~al. (2016).
\newblock The allelic landscape of human blood cell trait variation and links
  to common complex disease.
\newblock {\em Cell}, 167(5):1415--1429.

\bibitem[Balding and Nichols, 1995]{balding1995method}
Balding, D.~J. and Nichols, R.~A. (1995).
\newblock A method for quantifying differentiation between populations at
  multi-allelic loci and its implications for investigating identity and
  paternity.
\newblock {\em Genetica}, 96(1-2):3--12.

\bibitem[Benjamini et~al., 2006]{Benjamini2006}
Benjamini, Y., Krieger, A.~M., and Yekutieli, D. (2006).
\newblock {Adaptive linear step-up procedures that control the false discovery
  rate}.
\newblock {\em Biometrika}.

\bibitem[Consortium et~al., 2015]{phase3}
Consortium, . G.~P. et~al. (2015).
\newblock A global reference for human genetic variation.
\newblock {\em Nature}, 526(7571):68.

\bibitem[Ebejer et~al., 2013]{Ebejer2013}
Ebejer, J.~L., Duffy, D.~L., {Van Der Werf}, J., Wright, M.~J., Montgomery, G.,
  Gillespie, N.~A., Hickie, I.~B., Martin, N.~G., and Medland, S.~E. (2013).
\newblock {Genome-wide association study of inattention and
  hyperactivity-impulsivity measured as quantitative traits}.
\newblock {\em Twin Research and Human Genetics}.

\bibitem[Ellinghaus et~al., 2016]{ellinghaus2016analysis}
Ellinghaus, D., Jostins, L., Spain, S.~L., Cortes, A., Bethune, J., Han, B.,
  Park, Y.~R., Raychaudhuri, S., Pouget, J.~G., H{\"u}benthal, M., et~al.
  (2016).
\newblock Analysis of five chronic inflammatory diseases identifies 27 new
  associations and highlights disease-specific patterns at shared loci.
\newblock {\em Nature genetics}, 48(5):510.

\bibitem[Gao et~al., 2018]{gao2018genome}
Gao, X.~R., Huang, H., Nannini, D.~R., Fan, F., and Kim, H. (2018).
\newblock Genome-wide association analyses identify new loci influencing
  intraocular pressure.
\newblock {\em Human molecular genetics}, 27(12):2205--2213.

\bibitem[Harris, 2015]{Harris2015a}
Harris, K. (2015).
\newblock {Evidence for recent, population-specific evolution of the human
  mutation rate}.
\newblock {\em Proceedings of the National Academy of Sciences},
  112(11):3439--3444.

\bibitem[Harris and Pritchard, 2017]{Harris2017a}
Harris, K. and Pritchard, J.~K. (2017).
\newblock {Rapid evolution of the human mutation spectrum}.
\newblock {\em eLife}, 6.

\bibitem[Herold et~al., 2016]{herold2016family}
Herold, C., Hooli, B.~V., Mullin, K., Liu, T., Roehr, J.~T., Mattheisen, M.,
  Parrado, A.~R., Bertram, L., Lange, C., and Tanzi, R.~E. (2016).
\newblock Family-based association analyses of imputed genotypes reveal
  genome-wide significant association of alzheimer?s disease with osbpl6,
  ptprg, and pdcl3.
\newblock {\em Molecular psychiatry}, 21(11):1608.

\bibitem[{International HapMap Consortium}, 2005]{HapMap2005}
{International HapMap Consortium} (2005).
\newblock {A haplotype map of the human genome.}
\newblock {\em Nature}, 437(7063):1299--320.

\bibitem[Kraja et~al., 2011]{Kraja2011}
Kraja, A.~T., Vaidya, D., Pankow, J.~S., Goodarzi, M.~O., Assimes, T.~L.,
  Kullo, I.~J., Sovio, U., Mathias, R.~A., Sun, Y.~V., Franceschini, N.,
  Absher, D., Li, G., Zhang, Q., Feitosa, M.~F., Glazer, N.~L., Haritunians,
  T., Hartikainen, A.~L., Knowles, J.~W., North, K.~E., Iribarren, C., Kral,
  B., Yanek, L., O'Reilly, P.~F., McCarthy, M.~I., Jaquish, C., Couper, D.~J.,
  Chakravarti, A., Psaty, B.~M., Becker, L.~C., Province, M.~A., Boerwinkle,
  E., Quertermous, T., Palotie, L., Jarvelin, M.~R., Becker, D.~M., Kardia,
  S.~L., Rotter, J.~I., Chen, Y. D.~I., and Borecki, I.~B. (2011).
\newblock {A bivariate genome-wide approach to metabolic syndrome: STAMPEED
  Consortium}.
\newblock {\em Diabetes}.

\bibitem[Lan et~al., 2017]{Lan2017}
Lan, T., Lin, H., Zhu, W., Laurent, T. C. A.~M., Yang, M., Liu, X., Wang, J.,
  Wang, J., Yang, H., Xu, X., and Guo, X. (2017).
\newblock Deep whole-genome sequencing of 90 han chinese genomes.
\newblock {\em GigaScience}, 6(9):gix067.

\bibitem[Lek et~al., 2016]{Lek2016}
Lek, M., Karczewski, K.~J., Minikel, E.~V., Samocha, K.~E., Banks, E., Fennell,
  T., O'Donnell-Luria, A.~H., Ware, J.~S., Hill, A.~J., Cummings, B.~B.,
  Tukiainen, T., Birnbaum, D.~P., Kosmicki, J.~A., Duncan, L.~E., Estrada, K.,
  Zhao, F., Zou, J., Pierce-Hoffman, E., Berghout, J., Cooper, D.~N., Deflaux,
  N., DePristo, M., Do, R., Flannick, J., Fromer, M., Gauthier, L., Goldstein,
  J., Gupta, N., Howrigan, D., Kiezun, A., Kurki, M.~I., Moonshine, A.~L.,
  Natarajan, P., Orozco, L., Peloso, G.~M., Poplin, R., Rivas, M.~A.,
  Ruano-Rubio, V., Rose, S.~A., Ruderfer, D.~M., Shakir, K., Stenson, P.~D.,
  Stevens, C., Thomas, B.~P., Tiao, G., Tusie-Luna, M.~T., Weisburd, B., Won,
  H.~H., Yu, D., Altshuler, D.~M., Ardissino, D., Boehnke, M., Danesh, J.,
  Donnelly, S., Elosua, R., Florez, J.~C., Gabriel, S.~B., Getz, G., Glatt,
  S.~J., Hultman, C.~M., Kathiresan, S., Laakso, M., McCarroll, S., McCarthy,
  M.~I., McGovern, D., McPherson, R., Neale, B.~M., Palotie, A., Purcell,
  S.~M., Saleheen, D., Scharf, J.~M., Sklar, P., Sullivan, P.~F., Tuomilehto,
  J., Tsuang, M.~T., Watkins, H.~C., Wilson, J.~G., Daly, M.~J., and MacArthur,
  D.~G. (2016).
\newblock {Analysis of protein-coding genetic variation in 60,706 humans}.
\newblock {\em Nature}, 536(7616):285--291.

\bibitem[L{\'o}pez-Mej{\'\i}as et~al., 2017]{lopez2017genome}
L{\'o}pez-Mej{\'\i}as, R., Carmona, F.~D., Casta{\~n}eda, S., Genre, F.,
  Remuzgo-Mart{\'\i}nez, S., Sevilla-Perez, B., Ortego-Centeno, N., Llorca, J.,
  Ubilla, B., Mijares, V., et~al. (2017).
\newblock A genome-wide association study suggests the hla class ii region as
  the major susceptibility locus for iga vasculitis.
\newblock {\em Scientific reports}, 7(1):5088.

\bibitem[Lutz et~al., 2015]{lutz2015genome}
Lutz, S.~M., Cho, M.~H., Young, K., Hersh, C.~P., Castaldi, P.~J., McDonald,
  M.-L., Regan, E., Mattheisen, M., DeMeo, D.~L., Parker, M., et~al. (2015).
\newblock A genome-wide association study identifies risk loci for spirometric
  measures among smokers of european and african ancestry.
\newblock {\em BMC genetics}, 16(1):138.

\bibitem[Mafessoni et~al., 2018]{mafessoni2018turning}
Mafessoni, F., Prasad, R.~B., Groop, L., Hansson, O., and Pr{\"u}fer, K.
  (2018).
\newblock Turning vice into virtue: Using batch-effects to detect errors in
  large genomic data sets.
\newblock {\em Genome biology and evolution}, 10(10):2697--2708.

\bibitem[Mallick et~al., 2016]{Mallick2016}
Mallick, S., Li, H., Lipson, M., Mathieson, I., Gymrek, M., Racimo, F., Zhao,
  M., Chennagiri, N., Nordenfelt, S., Tandon, A., Skoglund, P., Lazaridis, I.,
  Sankararaman, S., Fu, Q., Rohland, N., Renaud, G., Erlich, Y., Willems, T.,
  Gallo, C., Spence, J.~P., Song, Y.~S., Poletti, G., Balloux, F., {Van Driem},
  G., {De Knijff}, P., Romero, I.~G., Jha, A.~R., Behar, D.~M., Bravi, C.~M.,
  Capelli, C., Hervig, T., Moreno-Estrada, A., Posukh, O.~L., Balanovska, E.,
  Balanovsky, O., Karachanak-Yankova, S., Sahakyan, H., Toncheva, D.,
  Yepiskoposyan, L., Tyler-Smith, C., Xue, Y., Abdullah, M.~S., Ruiz-Linares,
  A., Beall, C.~M., {Di Rienzo}, A., Jeong, C., Starikovskaya, E.~B., Metspalu,
  E., Parik, J., Villems, R., Henn, B.~M., Hodoglugil, U., Mahley, R.,
  Sajantila, A., Stamatoyannopoulos, G., Wee, J.~T., Khusainova, R.,
  Khusnutdinova, E., Litvinov, S., Ayodo, G., Comas, D., Hammer, M.~F.,
  Kivisild, T., Klitz, W., Winkler, C.~A., Labuda, D., Bamshad, M., Jorde,
  L.~B., Tishkoff, S.~A., Watkins, W.~S., Metspalu, M., Dryomov, S., Sukernik,
  R., Singh, L., Thangaraj, K., Pa{\"{a}}bo, S., Kelso, J., Patterson, N., and
  Reich, D. (2016).
\newblock {The Simons Genome Diversity Project: 300 genomes from 142 diverse
  populations}.
\newblock {\em Nature}, 538(7624):201--206.

\bibitem[Mandage et~al., 2017]{Mandage2017}
Mandage, R., Telford, M., Rodr{\'{i}}guez, J.~A., Farr{\'{e}}, X., Layouni, H.,
  Marigorta, U.~M., Cundiff, C., Heredia-Genestar, J.~M., Navarro, A., and
  Santpere, G. (2017).
\newblock {Genetic factors affecting EBV copy number in lymphoblastoid cell
  lines derived from the 1000 Genome Project samples}.
\newblock {\em PLoS ONE}.

\bibitem[Mathieson and Reich, 2017]{Mathieson2017a}
Mathieson, I. and Reich, D. (2017).
\newblock {Differences in the rare variant spectrum among human populations}.
\newblock {\em PLoS Genetics}, 13(2).

\bibitem[McCarthy et~al., 2016]{mccarthy2016reference}
McCarthy, S., Das, S., Kretzschmar, W., Delaneau, O., Wood, A.~R., Teumer, A.,
  Kang, H.~M., Fuchsberger, C., Danecek, P., Sharp, K., et~al. (2016).
\newblock A reference panel of 64,976 haplotypes for genotype imputation.
\newblock {\em Nature genetics}, 48(10):1279.

\bibitem[Minoche et~al., 2011]{Minoche2011}
Minoche, A.~E., Dohm, J.~C., and Himmelbauer, H. (2011).
\newblock {Evaluation of genomic high-throughput sequencing data generated on
  Illumina HiSeq and Genome Analyzer systems}.
\newblock {\em Genome Biology}, 12(11).

\bibitem[Nagy et~al., 2017]{nagy2017exploration}
Nagy, R., Boutin, T.~S., Marten, J., Huffman, J.~E., Kerr, S.~M., Campbell, A.,
  Evenden, L., Gibson, J., Amador, C., Howard, D.~M., et~al. (2017).
\newblock Exploration of haplotype research consortium imputation for
  genome-wide association studies in 20,032 generation scotland participants.
\newblock {\em Genome medicine}, 9(1):23.

\bibitem[Nishida et~al., 2018]{nishida2018key}
Nishida, N., Sugiyama, M., Sawai, H., Nishina, S., Sakai, A., Ohashi, J., Khor,
  S.-S., Kakisaka, K., Tsuchiura, T., Hino, K., et~al. (2018).
\newblock Key hla-drb1-dqb1 haplotypes and role of the btnl2 gene for response
  to a hepatitis b vaccine.
\newblock {\em Hepatology}, 68(3):848--858.

\bibitem[Park et~al., 2015]{park2015mercapturic}
Park, S.~L., Carmella, S.~G., Chen, M., Patel, Y., Stram, D.~O., Haiman, C.~A.,
  Le~Marchand, L., and Hecht, S.~S. (2015).
\newblock Mercapturic acids derived from the toxicants acrolein and
  crotonaldehyde in the urine of cigarette smokers from five ethnic groups with
  differing risks for lung cancer.
\newblock {\em PLoS One}, 10(6):e0124841.

\bibitem[Pfeifer et~al., 2002]{Pfeifer2002}
Pfeifer, G.~P., Denissenko, M.~F., Olivier, M., Tretyakova, N., Hecht, S.~S.,
  and Hainaut, P. (2002).
\newblock {Tobacco smoke carcinogens, DNA damage and p53 mutations in
  smoking-associated cancers}.
\newblock {\em Oncogene}, 21-48(6):7435--7451.

\bibitem[Pleasance et~al., 2010]{Pleasance2010}
Pleasance, E.~D., Stephens, P.~J., O'Meara, S., McBride, D.~J., Meynert, A.,
  Jones, D., Lin, M.~L., Beare, D., Lau, K.~W., Greenman, C., Varela, I.,
  Nik-Zainal, S., Davies, H.~R., Ord{\~{o}}ez, G.~R., Mudie, L.~J., Latimer,
  C., Edkins, S., Stebbings, L., Chen, L., Jia, M., Leroy, C., Marshall, J.,
  Menzies, A., Butler, A., Teague, J.~W., Mangion, J., Sun, Y.~A., McLaughlin,
  S.~F., Peckham, H.~E., Tsung, E.~F., Costa, G.~L., Lee, C.~C., Minna, J.~D.,
  Gazdar, A., Birney, E., Rhodes, M.~D., McKernan, K.~J., Stratton, M.~R.,
  Futreal, P.~A., and Campbell, P.~J. (2010).
\newblock {A small-cell lung cancer genome with complex signatures of tobacco
  exposure}.
\newblock {\em Nature}, 463(7278):184--190.

\bibitem[Pritchard et~al., 2000]{pritchard2000inference}
Pritchard, J.~K., Stephens, M., and Donnelly, P. (2000).
\newblock Inference of population structure using multilocus genotype data.
\newblock {\em Genetics}, 155(2):945--959.

\bibitem[Shiraishi et~al., 2015]{Shiraishi2015a}
Shiraishi, Y., Tremmel, G., Miyano, S., and Stephens, M. (2015).
\newblock {A Simple Model-Based Approach to Inferring and Visualizing Cancer
  Mutation Signatures}.
\newblock {\em PLoS Genetics}, 11(12).

\bibitem[Song et~al., 2015]{song2015testing}
Song, M., Hao, W., and Storey, J.~D. (2015).
\newblock Testing for genetic associations in arbitrarily structured
  populations.
\newblock {\em Nature genetics}, 47(5):550.

\bibitem[Spracklen et~al., 2017]{spracklen2017association}
Spracklen, C.~N., Chen, P., Kim, Y.~J., Wang, X., Cai, H., Li, S., Long, J.,
  Wu, Y., Wang, Y.~X., Takeuchi, F., et~al. (2017).
\newblock Association analyses of east asian individuals and trans-ancestry
  analyses with european individuals reveal new loci associated with
  cholesterol and triglyceride levels.
\newblock {\em Human molecular genetics}, 26(9):1770--1784.

\bibitem[Suhre et~al., 2017]{suhre2017connecting}
Suhre, K., Arnold, M., Bhagwat, A.~M., Cotton, R.~J., Engelke, R., Raffler, J.,
  Sarwath, H., Thareja, G., Wahl, A., DeLisle, R.~K., et~al. (2017).
\newblock Connecting genetic risk to disease end points through the human blood
  plasma proteome.
\newblock {\em Nature communications}, 8:14357.

\bibitem[Tian et~al., 2017]{tian2017genome}
Tian, C., Hromatka, B.~S., Kiefer, A.~K., Eriksson, N., Noble, S.~M., Tung,
  J.~Y., and Hinds, D.~A. (2017).
\newblock Genome-wide association and hla region fine-mapping studies identify
  susceptibility loci for multiple common infections.
\newblock {\em Nature communications}, 8(1):599.

\bibitem[van Dijk et~al., 2014]{VanDijk2014}
van Dijk, E.~L., Auger, H., Jaszczyszyn, Y., and Thermes, C. (2014).
\newblock {Ten years of next-generation sequencing technology}.
\newblock {\em Trends in Genetics}, 30(9):418--426.

\bibitem[Xu et~al., 2012]{xu2012genome}
Xu, J., Mo, Z., Ye, D., Wang, M., Liu, F., Jin, G., Xu, C., Wang, X., Shao, Q.,
  Chen, Z., et~al. (2012).
\newblock Genome-wide association study in chinese men identifies two new
  prostate cancer risk loci at 9q31. 2 and 19q13. 4.
\newblock {\em Nature genetics}, 44(11):1231.

\bibitem[Yucesoy et~al., 2015]{yucesoy2015genome}
Yucesoy, B., Kaufman, K.~M., Lummus, Z.~L., Weirauch, M.~T., Zhang, G.,
  Cartier, A., Boulet, L.-P., Sastre, J., Quirce, S., Tarlo, S.~M., et~al.
  (2015).
\newblock Genome-wide association study identifies novel loci associated with
  diisocyanate-induced occupational asthma.
\newblock {\em Toxicological Sciences}, 146(1):192--201.

\end{thebibliography}

@@

%\extrafloats{100}
